---
title: "Data Wrangling on one table"
author: "Gregory J. Matthews"
date: "8/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tidyverse grammar
- select(): Returns a sub set of columns (i.e. features or variables)
- filter(): Returns a subset of rows (i.e. observations)
- mutate(): add or modify existing columns
- arrange(): sort the rows
- summarize(): aggregate the data across rows (e.g., group it according to some criteria)

#Filter and Select
```{r}
library(tidyverse)
library(mdsr)
#Subset of rows
filter(CIACountries, pop > 1000000000)

#I can also do it like this
CIACountries %>% filter(pop > 1e9)

#In base R
CIACountries[CIACountries$pop > 1e9,]

#or
subset(CIACountries, pop > 1e9)

```


```{r}
#Subset of columns
select(CIACountries, country, pop, gdp)

#I can also do it like this
CIACountries %>% select(country, pop, gdp)

#In base R
CIACountries[, c("country", "pop", "gdp")]

#or
subset(CIACountries, select = c("country", "pop", "gdp"))

```


```{r}
#filter and select
filter(select(CIACountries, country, pop, gdp), pop > 1e9)

#Or we can do this with a pipeline, which is easier to read. 
#at the same time in the same line, order does not matter for these guys
CIACountries %>% select(country, pop, gdp) %>% filter(pop > 1e9)

#Or in base R
CIACountries[CIACountries$pop > 1e9,c("country", "pop", "gdp")]

#or this:
subset(CIACountries, pop > 1e9, select = c("country", "pop", "gdp"))

```
#Mutate and Rename
```{r}
#Mutate
#Population Density
my_CIACountries <- mutate(CIACountries, dens = pop/area)
my_CIACountries[1,]

my_CIACountries <- CIACountries %>% mutate(dens = pop/area)
my_CIACountries[1,]

#Base R
library(mdsr)
my_CIACountries$dens <- CIACountries$pop / CIACountries$area

#rename
my_CIACountries <- CIACountries %>% rename(education = educ)
my_CIACountries[1,]
```

#arrange

```{r}
#Sorting with arrange
#Defaults to ascending
CIACountries_sort <- arrange(CIACountries, gdp)
CIACountries_sort[1:5,]

#Sort descending
CIACountries_sort <- arrange(CIACountries, desc(gdp))
CIACountries_sort[1:5,]

CIACountries_sort <- CIACountries %>% arrange(desc(gdp))
CIACountries_sort[1:5,]

#Base R with order
CIACountries_sort <- CIACountries[order(CIACountries$gdp),]
CIACountries_sort[1:5,]

CIACountries_sort <- CIACountries[order(-CIACountries$gdp),]
CIACountries_sort[1:5,]
```

#Summarize and Group_by
```{r}
#What is the average area by high vs low education countries?
my_CIACountries <- CIACountries %>% mutate(high_educ = educ > 4.5) %>% group_by(high_educ) %>% summarize(mn_area = mean(area))
my_CIACountries

#Base R
my_CIACountries <- CIACountries
my_CIACountries$high_educ <- CIACountries$educ > 4.5
tapply(my_CIACountries$area,my_CIACountries$high_educ, mean)

```


##In class assignment: What first letter has the largest population? 
```{r}
#My_CIACountries <- CIACountries %>% mutate(letter = substring(country,1,1))
# adds the variable for letter. fletter is a better name! 
#'3 more until answer'
#population_and_letter <- subset(My_CIACountries, select = c("country", "pop", "letter"))
#add_by_letter <- population_and_letter %>% group_by(letter) %>% #summarise(sum(pop))
#add_by_letter
#arrange in descending order and call the first row
#My_CIACountries <- CIACountries %>% mutate(letter = substring(country,1,1)) %>%  #group_by(letter) %>% summarise(totalpop = sum(pop))

CIACountries %>% 
  mutate(letter = substring(country,1,1)) %>%  
  group_by(letter) %>% 
  summarise(totalpop = sum(pop)) %>%
  arrange(-totalpop) %>%
  head(5)


```


#What proportion of countries produce any oil stratified by net users class? 

```{r}
#stratify is summarise by group, the groups are "net users" class
#proportion how??? use mutate at least
#CIACountries %>% 
#  summarise(oil_prod, net_users) #%>%
#  mutate(proportion = CIACountries$as.integer(oil_prod) / #CIACountries$as.integer(net_users))

#think of what the answer will look like. net users table, with number between 0 and 1, which is proportion of those that produce oil. new variable using mutate
CIACountries %>% 
  mutate(anyoil = (oil_prod > 0) + 0) %>% # boolean trues and falses
  group_by(net_users) %>% #doesnt change just preps for summary
  summarise(oil_prop = mean(anyoil, na.rm = TRUE)) 
  #can't compute mean with NAs there, so use na.rm aka remove
  #NA omit function will remove whole record if has one NA.
#"This shows that those who use more internet come from countries with more money and thus it is likely also oil production"   

```



  





